Phenotypic evolution is more constrained in simpler food webs

### 2019-01-09 manuscript.tex
update fig. 3 with arrows and contours

\emph{Keywords}: fitness landscape, host-parasitoid, natural selection,
G-matrix, community context, ecological networks.

@@

@@

Global change is simplifying the structure of ecological communities,
which may influence evolutionary constraints in a number of ways. For
example, in a multitrophic community or food web, if different consumers
impose directional selection on different traits of a shared resource,
then more complex food webs may constrain evolution by increasing the
number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more complex food webs. To examine these
different possibilities (among others), we conducted a field experiment
that manipulated the complexity of trophic interactions (hereafter
food-web complexity) associated with an abundant insect herbivore
(\emph{Iteomyia salicisverruca}; Family Cecidomyiidae)(fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortaltiy from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web complexity by either excluding the guild of larval
parasitoids (simple food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (complex food web;
fig. \ref{fig:Conceptual}). We then applied modern statistical methods
to quantify the effect of food-web complexity on the slope and curvature
of the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work \citep{Barbour2016}
and others work with gall midges \citep{Weis1983, Heath2018}. First, we
measured gall diameter as the size of each gall chamber to the nearest
0.01 mm at its maximum diameter (perpendicular to the direction of plant
tissue growth). Our previous work has shown that larger galls are
associated with higher survival \citep{Barbour2016}. Second, we measured
clutch size by counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple of features of our study system suggest that larval densities may
be a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

@@

The indirect effects of selection on genetic constraints is governed by
the curvature of the adaptive landscape, which in our study corresponds
to:

@@

All analyses and visualizations were conducted in R \citep{R2018}.
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population \citep{Arnold2001}.
We also plot mean larval survival on a natural log scale to accurately
reflect the shape of the adaptive landscape \citep{Arnold2003}.

@@

\subsection{Food-web complexity reduces selective
constraints}\label{food-web-complexity-reduces-selective-constraints}

\indent We found that more gall midge traits experienced directional
selection in the simple (3 of 3) vs.~complex food webs (1 of 3)(table
\ref{Table:Gradients}). For example, we observed directional selection
for smaller clutch sizes in the simple food web, but there was no
evidence of selection acting on this trait in the complex food web
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This
absence of selection appeared to be a result of conflicting selection
pressures imposed by each guild of parasitoids (fig.
\ref{fig:UV_Landscape}C). Specifically, when we subset our data to focus
on differences between parasitoid guilds, we found that larval
parasitoids actually impose directional selection for larger clutch
sizes (\(\beta_{\text{Clutch}}\)= 0.13 {[}0.03, 0.23{]}). In the simple
food web, we also observed clear evidence of directional selection for
midges to avoid ovipositing on plants with high densities of
conspecifics (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}B); however, this relationship was weaker in
complex food webs (table \ref{Table:Gradients}). This was likely a
result of larval parasitoids imposing greater mortality on egg
parasitoids at high gall midge densities (see Selection on the extended
phenotype of egg parasitoids section), and thus a less than additive
effect on gall midges. Chamber diameter experienced positive directional
selection in both food-web treatments, but selection was 65\% higher in
complex food webs (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}). This was not due to any difference between egg
and larval parasitoids (fig. \ref{fig:UV_Landscape}), but was due to the
lower average survival in complex food webs (contrast -0.26 {[}-0.12,
-0.42{]}).

\subsection{Food-web complexity reduces genetic
constraints}\label{food-web-complexity-reduces-genetic-constraints}

\indent The curvature of the adaptive landscape indirectly affects
genetic constraints and is influenced by directional, quadratic, and
correlational selection gradients. Our food-web treatment did not alter
correlational selection for any combination of traits (table
\ref{Table:Gradients}). Similarly, our food-web treatment did not
influence quadratic selection on either chamber diameter or clutch size
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}A,C). In
contrast, our food-web treatment did alter quadratic selection acting on
oviposition preference (table \ref{Table:Gradients}). The negative
relationship between oviposition preference and larval survival dampened
at high densities in complex, but not simple food webs (fig.
\ref{fig:UV_Landscape}). This dampened relationship was partly due to a
trend for nonlinear selection imposed by larval parasitoids
(\(\gamma_{\text{Pref:Pref}}\)= 0.18 {[}-0.01, 0.42{]}), but was also
magnified by the lower average survival in complex food webs.

@@

\begin{enumerate}

 \item Summary of key results
    \begin{enumerate}
    \item Gall midges experience more selective and constraints in simple food webs.
    \end{enumerate}

 \item Natural history and comparison to other studies of natural selection in galling insects.
    \begin{enumerate}
    \item \textbf{Important Studies for Comparison}: @Abrahamson1997; @Weis1983; @Heath2018; Start, Weis, and Gilbert In Press Am. Nat; @Start2016; @Start2018urbanization; Peter Price's work; Timothy Craig's work; Cheryl Briggs work.
    \item \textbf{Topics to discuss}: Persistent pattern in selection for larger galls; Contrasting effects of egg and larval parasitoids on clutch size; Nonlinear selection at high gall densities is likely reflective of the poorer searching ability of larval vs. egg parasitoids (and thus it saturates more quickly).  
    \end{enumerate}

 \item Selective constraints:
    \begin{enumerate}
    \item \textbf{Key Result}: More traits under directional selection in simple (3 of 3) vs. complex (1 of 3) food webs. Need to be wary of interpreting no selection on oviposition preference in complex food webs though. Even if there is selection on oviposition preference, this does not change our qualitative result.
    \item \textbf{Implications}: The trajectory of phenotypic evolution is less constrained in complex vs. simple food webs. In other words, there are more potential adaptive peaks in complex food webs. 
    \end{enumerate}

 \item Indirect effects of selection on genetic constraints: 
    \begin{enumerate}
    \item \textbf{Key Result}: More traits experience a decrease in additive genetic variance and more trait combinations experience selection for trait integration in simple food webs. 
    \item \textbf{Implications}: The structure of the G-matrix is under less influence of selection in complex vs. simple food webs. Thus the trajectory of phenotypic evolution may be dictated more by genetic rather than selective constraints. This is in line with empirical evidence (from complex natural systems) that evolution tends to proceed along genetic lines of least resistance [@Schluter1996].
    \end{enumerate}

 \item Indirect evolutionary effects:
    \begin{enumerate}
    \item \textbf{Key Result}: Evidence of selection on the extended phenotype of egg parasitoids.
    \item \textbf{Implications}: Hints at a dynamic adaptive landscape and potential indirect evolutionary effects in complex food webs.
    \end{enumerate}

 \item Caveats: 
    \begin{enumerate} 
    \item \textbf{Key Caveat}: Experimental artifact of lower larval survival in complex food-web treatment.
    \item \textbf{Implications}: Artificially increased the magnitude of selection in complex food webs. Note however, that this does not qualitatively change any of our results. In fact, we are actually underestimating the constraints imposed by our simple food-web treatment. This is because the egg parasitoids would be released from their intraguild predator, and thus could increase in abundance over future generations and impose stronger selection on gall midges. 
    \end{enumerate}

\item Conclusions

\end{enumerate}

@@

@@

\section{Figure Legends}\label{figure-legends}

\textbf{Figure \ref{fig:Conceptual}}: Experimental manipulation of
food-web complexity associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. In the
complex food-web treatment (A), we allowed the full suite of egg (blue
rectangle) and larval (yellow rectangle) parasitoids to impose
selection. In the simple food-web treatment (B), we used mesh bags to
exclude the guild of larval parasitoids, only allowing the egg
parasitoid (\emph{Platygaster} sp.) to impose selection. Black arrows
denote the flow of energy in this network of trophic interactions.
Larval parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).

\textbf{Figure \ref{fig:UV_Landscape}}: Adaptive landscape of gall midge
phenotypes in complex vs.~simple food webs. Each panel corresponds to a
different phenotypic trait: chamber diameter (A); oviposition preference
(B); and clutch size (C). Solid lines represent selection experienced in
complex (orange) and simple (blue) food webs. Transparent lines
represent bootstrapped replicates to show the uncertainty in selection.
For clarity, we only display 100 bootstraps even though inferences are
based on 1,000 replicates.Note that mean larval survival is plotted on a
natural log scale to reflect the mathematical defintion of the adaptive
landscape.

\textbf{Figure \ref{fig:MV_Landscape}}: Two dimensional view of adaptive
landscapes of gall midge phenotypes in complex vs.~simple food webs.
Each panel corresponds to a different combination of phenotypic traits:
clutch size and chamber diameter (A); clutch size and oviposition
preference (B); oviposition preference and chamber diameter (C). Note
the flat landscape in mean larval survival for clutch size and
oviposition preference in more complex food webs, whereas all three
traits shape the adaptive landscape in simpler food webs. This indicates
that selection favors trait integration for each combination of traits
in simpler food webs, whereas there is no strong evidence for trait
integration in more complex food webs. Note that mean larval survival is
plotted on a natural log scale to reflect the mathematical defintion of
the adaptive landscape.

\textbf{Figure \ref{fig:EggPtoid_Selection}}: Selection imposed by
larval parasitoids on the extended phenotype of egg parasitoids
(\emph{Platygaster} sp.). The solid line represents the average
difference in the probability of observing the egg parasitoid in complex
vs.~simple food webs as a function of gall midge oviposition preference.
Transparent lines represent bootstrapped replicates to show the
uncertainty in selection. For clarity, we only display 100 bootstraps
even though inferences are based on 1,000 replicates. The decrease in
the probability of observing egg parasitoids at high gall-midge
densities indicates the increasing impact of larval parasitoids.

\begin{figure}
\centering
\includegraphics{analyses/complex_simple_foodwebs.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
complexity associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. In the
complex food-web treatment (A), we allowed the full suite of egg (blue
rectangle) and larval (yellow rectangle) parasitoids to impose
selection. In the simple food-web treatment (B), we used mesh bags to
exclude the guild of larval parasitoids, only allowing the egg
parasitoid (\emph{Platygaster} sp.) to impose selection. Black arrows
denote the flow of energy in this network of trophic interactions.
Larval parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in complex vs.~simple food webs.
Each panel corresponds to a different combination of phenotypic traits:
clutch size and chamber diameter (A); clutch size and oviposition
preference (B); oviposition preference and chamber diameter (C).
Selection favors trait integration for each combination of traits in
simple food webs, whereas there is no strong evidence for trait
integration in complex food webs. Note that mean larval survival is
plotted on a natural log scale to accurately reflect the adaptive
landscape.}
\end{figure}

@@

\renewcommand\refname{Appendix}
\bibliography{references}

>>>>

\emph{Keywords}: fitness landscape, host-parasitoid, natural selection,
G-matrix, community context, ecological networks, evolutionary ecology,
eco-evolutionary dynamics.

@@

\begin{quote}
``What escapes the eye, however, is a much more insidious kind of
extinction: the extinction of ecological interactions.''
\citet{Janzen1974}
\end{quote}

@@

Global change is simplifying the structure of ecological communities,
which may influence evolutionary constraints in a number of ways. For
example, in a multitrophic community or food web, if different consumers
impose directional selection on different traits of a shared resource,
then more complex food webs may constrain evolution by increasing the
number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more complex food webs. To examine these
different possibilities (among others), we conducted a field experiment
that manipulated the complexity of trophic interactions (hereafter
food-web complexity) associated with an abundant insect herbivore
(\emph{Iteomyia salicisverruca}; Family Cecidomyiidae)(fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortality from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web complexity by either excluding the guild of larval
parasitoids (simple food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (complex food web;
fig. \ref{fig:Conceptual}). We then applied modern statistical methods
to quantify the effect of food-web complexity on the slope and curvature
of the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work \citep{Barbour2016}
and others work with gall midges \citep{Weis1983, Heath2018}. First, we
measured gall diameter as the size of each gall chamber to the nearest
0.01 mm at its maximum diameter (perpendicular to the direction of plant
tissue growth). Our previous work has shown that larger galls are
associated with higher survival \citep{Barbour2016}. Second, we measured
clutch size by counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this by randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple of features of our study system suggest that larval densities may
be a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

@@

The indirect effects of selection on genetic constraints is governed by
the curvature of the adaptive landscape
(\(\text{C}=\gamma - \beta \beta^\text{T}\)), which in our study
corresponds to:

@@

All analyses and visualizations were conducted in R \citep{R2018}.
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population \citep{Arnold2001}.
We also plot mean larval survival on a natural log scale to accurately
reflect the mathematical definition of the adaptive landscape
\citep{Arnold2003}.

@@

\subsection{Simpler food webs increase selective
constraints}\label{simpler-food-webs-increase-selective-constraints}

\indent We found that more gall midge traits experienced directional
selection in the simple (3 of 3) vs.~complex food-web treatment (1 of
3)(table \ref{Table:Gradients}). For example, we observed directional
selection for smaller clutch sizes in the simple food web, but there was
no evidence of selection acting on this trait in the complex food web
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This
absence of selection appeared to be a result of conflicting selection
pressures imposed by each guild of parasitoids (fig.
\ref{fig:UV_Landscape}). Specifically, when we subset our data to focus
on differences between parasitoid guilds, we found that larval
parasitoids actually impose directional selection for larger clutch
sizes (\(\beta_{\text{Clutch}}\)= 0.13 {[}0.03, 0.23{]}). In the simple
food web, we also observed clear evidence of directional selection for
midges to avoid ovipositing on plants with high densities of
conspecifics (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}B); however, this relationship was weaker in
complex food webs (table \ref{Table:Gradients}). This was likely a
result of larval parasitoids imposing greater mortality on egg
parasitoids at high gall midge densities (see Selection on the extended
phenotype of egg parasitoids section), and thus a less than additive
effect on gall midges. Chamber diameter experienced positive directional
selection in both food-web treatments, but selection was 65\% higher in
complex food webs (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}). This was not due to any difference between egg
and larval parasitoids (fig. \ref{fig:UV_Landscape}), but was due to the
lower average survival in complex food webs (contrast -0.26 {[}-0.12,
-0.42{]}).

\subsection{Simpler food webs increase genetic
constraints}\label{simpler-food-webs-increase-genetic-constraints}

\indent The curvature of the adaptive landscape indirectly affects
genetic constraints and is influenced by directional, quadratic, and
correlational selection gradients. Our food-web treatment did not alter
correlational selection for any combination of traits (table
\ref{Table:Gradients}). Similarly, our food-web treatment did not
influence quadratic selection on either chamber diameter or clutch size
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}A,C). In
contrast, our food-web treatment did alter quadratic selection acting on
oviposition preference (table \ref{Table:Gradients}). The negative
relationship between oviposition preference and larval survival dampened
at high densities in complex, but not simple food webs (fig.
\ref{fig:UV_Landscape}B). This dampened relationship was partly due to a
trend for nonlinear selection imposed by larval parasitoids
(\(\gamma_{\text{Pref:Pref}}\)= 0.18 {[}-0.01, 0.42{]}), but was also
magnified by the lower average survival in complex food webs.

@@

Our key finding was that phenotypic evolution in gall midges was more
constrained in simpler food webs. More traits contributed to the slope
of the adaptive landscape in the simpler food web, suggesting that the
trajectory of phenotypic evolution is more constrained. Similarly, the
simple food-web treatment indirectly increased the number of genetic
constraints, which could act to constrain the gall midge's adaptive
potential in the face of novel selection pressures. We also found
evidence of indirect selection pressures in more complex food webs,
suggesting that the adaptive landscape may be more dynamic in complex
food webs. Taken together, our study provides experimental evidence from
the field that the simplification of ecological communities may
constrain the adaptive potential of remaining populations.

All three phenotypic traits we examined experienced directional
selection in the simpler food web, indicating that there is an optimal
phenotype that maximizes larval survival (i.e.~large chamber diameter,
small clutch size, and avoidance of conspecifics). In contrast, we did
not observe selective constraints on clutch size and oviposition
preference in the more complex food web. This suggests that there are
many optimal phenotypes (adaptive peaks), or in this case, a flat
adaptive landscape where there are no fitness consequences for
phenotypic change in these traits. This also implies that as selective
constraints dampen in more complex food webs, then the trajectory of
evolution becomes more determined by genetic constraints. This is made
clear if we inspect the `Lande equation', \(\Delta z = \text{G} \beta\),
where G corresponds to the population's G-matrix. As the components of
\(\beta\) go to zero, then \(\Delta z\) is more influenced by the
G-matrix. Interestingly, this corresponds to \citet{Schluter1996}`s
hypothesis that phenotypic evolution often follows 'genetic lines of
least resistance'. \citet{Schluter1996} found support for this
hypothesis from data on natural populations of several vertebrate
species, including threespine sticklebacks, a few species of songbirds,
and mice from the genus \emph{Peromyscus}. All of these species occupy
intermediate trophic levels and are likely embedded in complex food
webs, which is consistent with our suggestion that genetic constraints
may have a stronger influence in more complex food webs.

We also found evidence for more genetic constraints in simpler food webs
due to indirect effects of selection on the population's G-matrix. The
ability of a population to adapt to novel selection pressures
(evolvability) is largely governed by the structure of its G-matrix
\citep{Hansen2008}. When selection favors genetic covariance between
traits (positive or negative), this results in less autonomy of
evolutionary responses to changing environments. Similarly, decreases in
genetic variance constrain potential for the trait itself to evolve.
Together, this suggests that simpler food webs may decrease the
evolutionary potential of populations by indirectly selecting for
decreases in genetic variance in multiple traits and favoring trait
integration. Current theory often assumes genetic variances and
covariances remain constant over time rather than dynamically changing
with the community context \citep{McPeek2017, Guimaraes2017}. Our
empirical results highlight the need to explore the evolutionary
consequences of not only direct effects of selection, but indirect
effects on genetic constraints that are shaped by the community context.

The generality of our results likely depends on the relative abundance
and functional differences between consumers in a community. In the
simplest case, when consumers do not differ from each other, then we do
not expect changes in food-web complexity to modify selective
constraints. Also, many consumers may be at too low of abundances to
impose selection on their resources. Rank abundance curves
\citep{Preston1948} and the disproportionate number of weak interactions
in diverse communities \citep{Paine1992} support this notion. When
consumers are abundant though, the effect of food-web complexity will
depend on whether different species impose conflicting selection
pressures or select for distinct traits. For example, parisitoids and
birds impose conflicting selection pressures on the size of galls
induced by the fly \emph{Eurosta solidaginis}
\citep{Weis1985, Abrahamson1997, Start2016}. Recent studies in this
system have shown that decreases in the relative abundance of birds, due
to either small patch sizes \citep{Start2016} or proximity to urban
areas \citep{Start2018} causes a shift from neutral to directional
selection on gall size. On the other hand, different consumers may
impose selection on different traits, favoring trait integration and
increasing genetic constraints. Examples of this include strong genetic
covariances in plant resistance to different insect herbivores
\citep{Maddox1990, Wise2007, Wise2013}, although there are also examples
where these covariances are weak \citep{Roche1997, Barbour2015}, or vary
from year-to-year \citep{Johnson2007}. We suggest that gaining
predictive insight to the effects of food-web disassembly requires an
understanding of the mechanisms governing the assembly of trophic
interactions.

Our results suggest that the simplification of ecological communities
may not only directly affect connected species, but also result in
indirect evolutionary effects. In our study, this indirect effect arises
from egg parasitoids being released from intraguild predation in simpler
food webs. This release occurs more on trees with high larval densities,
which could intensify selection on gall midge oviposition preference. A
growing number of experiments over the past two decades have
demonstrated the presence and potential importance of indirect
evolutionary effects that emerge in a community context
\citep{Juenger1998, Stinchcombe2001, Lankau2007, Walsh2008, Walsh2010, terHorst2010, Sahli2011, Lau2012, terHorst2015, Schiestl2018}.
If indirect evolutionary effects are common \citep{Walsh2013}, then
predicting evolutionary trajectories resulting from the simplification
of food webs will require evolutionary studies to explicitly account for
the ecological networks that species are embedded in.

There are a growing number of theoretical studies on adaptation to
directional environmental change that incorporate species interactions
\citep[e.g.][]{deMazancourt2008, Johansson2008, Norberg2012, Osmond2017PredatorsHelpPrey}.
This work has given insight to the mechanisms by which pairwise
interactions modify the adaptive potential of species. Our study hints
at a novel mechanism, whereby food-web complexity flattens the adaptive
landscape, thus facilitating future adaptation by allowing genetic and
phenotypic variation to persist. This mechanism only emerges once we
move beyond pairwise interactions to consider selection on multiple
traits in a community context. However, a study on competition has
highlighted that we may expect the opposite effect of species diversity
in competitive communities \citep{deMazancourt2008}. This negative
effect of species diversity on adaptation occurs because there is a
greater chance that the optimal phenotype is already occupied by a
competitor in a more diverse community. More work is needed to examine
how the distribution of different interaction types affects adaptation
to environmental change in species-rich communities.

Our study gives insight to how local extinctions, and concomitant
decreases in food-web complexity, increase evolutionary constraints.
This hints at a potential insidious effect of local extinctions that
compromises the robustness of remaining populations to future
environmental change. Our work also highlights some key challenges for
predicting phenotypic evolution in the face of global change. First, the
simplification of ecological communities may actually reduce the
predictability of phenotypic evolution, because knowledge is required of
both selective and genetic constraints, rather than potentially just
genetic constraints in more complex systems. Second, many theoretical
models of eco-evolutionary dynamics focus on phenotypic change in a
single trait, yet our results highlight that the number of traits under
selection may change with the community context. Importantly, while the
net effects of selection in complex communities may be dampened, we
still found that different species/guilds imposed different selection
pressures. Knowing these hidden selection pressures is critical for
prediction, because the trajectory of evolution will depend on the
nature of change in the community context. We expect that a continued
integration of adaptive landscapes and ecological networks will enhance
our ability to predict the evolutionary consequences of changes in
ecological communities.

@@

We thank the staff of Humboldt Bay National Wildlife Refuge (U.S. Fish
and Wildlife Service) for facilitating experimental logistics. For
assistance with fieldwork, we thank Ruthie Espanol and Andrew MacDonald.
For funding support, we thank the University of British Columbia (James
Robert Thompson Fellowship and Four-Year Fellowship to M.A.~Barbour),
NSERC (Discovery grant to Greg Crutsinger), and the Swiss National
Science Foundation (grant 31003A\_160671 to J. Bascompte).

@@

\begin{figure}
\centering
\includegraphics{analyses/complex_simple_foodwebs_v2.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
complexity associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the complex food-web treatment, we allowed the full
suite of egg and larval parasitoids to impose selection. In the simple
food-web treatment, we used mesh bags to exclude the guild of larval
parasitoids, only allowing the egg parasitoid (\emph{Platygaster} sp.)
to impose selection. Larval parasitoids include the following species
(from left to right): \emph{Mesopolobus} sp. (Family: Pteromalidae);
\emph{Tetrastichus} sp. (Family: Eulophidae); and \emph{Torymus} sp.
(Family: Torymidae).}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in complex vs.~simple food webs.
Each panel corresponds to a different combination of phenotypic traits:
clutch size and chamber diameter (A); clutch size and oviposition
preference (B); oviposition preference and chamber diameter (C). Arrows
represent mean estimates of directional selection gradients, while
contours represent predicted larval survival of the mean phenotype
(i.e., all traits = 0) in each food-web treatment. Notice that arrows
point more toward a corner of the adaptive landscape for each
combination of traits in the simple vs.~complex food-web treatment. This
indicates that trait integration (covariance) is more strongly favored
in simpler vs.~more complex food webs. Note that mean larval survival is
plotted on a natural log scale to reflect the mathematical definition of
the adaptive landscape.}
\end{figure}

@@

\bibliography{references}
